---
title: "UCLA Already Has a Research Data Roadmap"
subtitle: "What the RDWG Found — and What It Means for OneIT"
author: "Tim Dennis, UCLA Library Data Science Center"
date: "March 2026"
format:
  revealjs:
    theme: [simple, ucla-theme.scss]
    logo: assets/logos/logo_UCLA_blue.svg
    slide-number: true
    footer: "One IT Research Working Group — March 2026"
    fig-align: center
    echo: false
    warning: false
    message: false
title-slide-attributes:
  data-background-color: "#003B5C"
execute:
  echo: false
  warning: false
---

```{r setup}
library(ggplot2)
library(dplyr)
library(scales)

# UCLA brand colors
ucla_blue      <- "#2774AE"
ucla_gold      <- "#FFD100"
ucla_dark_blue <- "#003B5C"
ucla_light_blue <- "#8BB8E8"
```

## Our Charge

The One IT Research WG is tasked with:

- Assessing critical research digital experiences
- Identifying gaps and investment priorities
- Evaluating campus IT roles
- Recommending improvements

**UCLA's Research Data Working Group spent 18 months doing exactly this.**

Their findings are directly relevant — and already available to us.

---

## What Was the RDWG?

**Charged in early 2021 by five executive sponsors:**

- Roger Wakimoto — Vice Chancellor for Research
- Ginny Steel — University Librarian
- Lucy Avetisyan — AVC & CIO
- Jim Davis — Vice Provost for IT
- Ellen Pollack — CIO, UCLA Health Sciences

**18 months of work:** campus-wide data gathering, faculty input, peer benchmarking across 13 service categories

**Output:** *A Call to Action* — 66-page roadmap, December 2022

---

## UCLA's NSF Per-Award Gap

```{r nsf-chart}
#| fig-height: 4.2

nsf <- data.frame(
  institution = c("UC Santa Barbara", "UC San Diego", "Stanford",
                  "UC Berkeley", "UC Irvine", "UCLA", "UC Davis"),
  award_avg = c(427708, 411669, 393089, 376572, 311264, 307474, 302687)
) |>
  mutate(
    institution = factor(institution, levels = rev(institution)),
    is_ucla = institution == "UCLA",
    label = paste0("$", comma(round(award_avg / 1000), accuracy = 1), "K")
  )

ggplot(nsf, aes(x = award_avg, y = institution, fill = is_ucla)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = label), hjust = -0.15, size = 3.5, color = "#1a1a1a") +
  scale_fill_manual(values = c("FALSE" = ucla_light_blue, "TRUE" = ucla_gold),
                    guide = "none") +
  scale_x_continuous(labels = dollar_format(scale = 1/1000, suffix = "K"),
                     limits = c(0, 510000), expand = c(0, 0)) +
  labs(
    title = "NSF 5-Year Average Per-Award Value",
    subtitle = "UCLA ranks 6th of 7 UC peers — a 15–30% gap below UCSB and UCSD",
    x = NULL, y = NULL,
    caption = "Source: NSF Budget Internet Information System (RDWG, 2022)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(color = ucla_dark_blue, face = "bold", size = 14),
    plot.subtitle = element_text(color = "#444444", size = 11),
    plot.caption  = element_text(color = "#888888", size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y = element_text(
      color = ifelse(levels(nsf$institution) == "UCLA", ucla_dark_blue, "#1a1a1a"),
      face  = ifelse(levels(nsf$institution) == "UCLA", "bold", "plain"),
      size  = 12
    )
  )
```

---

## The Core Problem: Coordination, Not Capability

UCLA fell behind peers in **8 of 13** research data service categories:

```{r benchmarking-chart}
#| fig-height: 3.5

bench <- data.frame(
  status = c("Behind peers", "At par or ahead"),
  count  = c(8, 5)
) |>
  mutate(
    pct   = count / sum(count),
    label = paste0(count, " of 13\ncategories"),
    ypos  = cumsum(pct) - pct / 2
  )

ggplot(bench, aes(x = 2, y = pct, fill = status)) +
  geom_col(width = 1, color = "white", linewidth = 1.2) +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5),
            size = 4.5, color = "white", fontface = "bold") +
  coord_polar(theta = "y", start = 0) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = c("Behind peers" = ucla_dark_blue,
                               "At par or ahead" = ucla_light_blue)) +
  labs(
    title    = "Peer Benchmarking: 13 Research Data Service Categories",
    subtitle = "Lowest score among peers: research data services program\nwith a website and service catalog",
    fill     = NULL,
    caption  = "Source: RDWG peer benchmarking (Research on Data Services II, 2022)"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(color = ucla_dark_blue, face = "bold",
                                 size = 13, hjust = 0.5),
    plot.subtitle = element_text(color = "#444444", size = 10, hjust = 0.5),
    plot.caption  = element_text(color = "#888888", size = 8, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 11)
  )
```

---

## The Compliance Environment

The stakes have risen significantly since 2022:

- **NIH Data Management & Sharing Policy** — in effect January 2023; every proposal requires a funded data management plan
- **OSTP public access mandate** — in effect December 2025; all federally-funded research publicly available at no cost
- **UC Research Data Policy** — VCR responsible for campus compliance
- **Export controls** — tightened significantly on AI and dual-use research

**UCLA lacks the staff and infrastructure to support compliance at the scale of a $1.7B research enterprise.**

---

## What Peer Institutions Do

> *"Investing in central services enhances distributed efforts — it does not replace them."*

```{mermaid}
mindmap
  root((Layered Research IT))
    Centralize the Backbone
      Network and IAM/SSO
      Shared storage baseline
      Cybersecurity
      Cloud procurement
      Compliance coordination
    Preserve Domain Expertise
      Research computing
      HPC systems
      Data services consulting
      Research software engineering
      Discipline-specific platforms
```

---

## What to Centralize — What to Preserve

:::: {.columns}

::: {.column}
**Centralize the backbone:**

- Network infrastructure, IAM/SSO, cybersecurity
- Shared storage (baseline for all faculty)
- Cloud procurement and cost coordination
- Compliance guidance and service catalog
:::

::: {.column}
**Preserve domain expertise:**

- Research computing (PI-facing, domain-specific)
- HPC and specialized research systems
- Data services consulting (FAIR, DMPs, repositories)
- **UCLA Redivis** — sensitive data, already built
:::

::::

> *"Centralize the backbone — not the craft."*

This model is called **Academic Divisional Computing (ADC)** — documented across UC institutions and peer research universities as the approach that preserves domain expertise while achieving back-end efficiencies.

---

## The Four RDWG Recommendations

**1. Central coordination unit** *(proposed: ORDR)*
Clear ownership of research data campus-wide — service catalog, compliance lead, accountability. *Governance: VCR.*

**2. Shared research storage** *(proposed: Bruin Research Data Hub)*
Baseline allocation for every faculty member at no cost. On-prem + cloud hybrid.

**3. Expanded research data services**
Data science consulting, digital asset management, researcher education. Built on existing Library/DSC capacity.

**4. Expanded research infrastructure**
HPC expansion, P3/P4 compliant environment for restricted data, Cloud for Research program.

---

## Investment Scale

```{r investment-chart}
#| fig-height: 3.8

invest <- data.frame(
  component = c("Coordination\n(ORDR)",
                "Shared Storage\n(Bruin Hub)",
                "Research Data\nServices",
                "Research\nInfrastructure"),
  full = c(9.2, 25.3, 11.1, 28.6),
  mvp  = c(7.1, 23.0,  5.5, 14.2)
) |>
  tidyr::pivot_longer(c(full, mvp), names_to = "scenario", values_to = "cost") |>
  mutate(
    component = factor(component, levels = rev(unique(component))),
    scenario  = factor(scenario, levels = c("full", "mvp"),
                       labels = c("Full Implementation ($74.2M)", "MVP ($49.8M)"))
  )

ggplot(invest, aes(x = cost, y = component, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0("$", cost, "M")),
            position = position_dodge(width = 0.7),
            hjust = -0.1, size = 3.2, color = "#1a1a1a") +
  scale_fill_manual(values = c("Full Implementation ($74.2M)" = ucla_dark_blue,
                               "MVP ($49.8M)" = ucla_light_blue)) +
  scale_x_continuous(labels = dollar_format(suffix = "M"),
                     limits = c(0, 36), expand = c(0, 0)) +
  labs(
    title   = "6-Year Investment Range (2022 RDWG Estimates)",
    x = "6-Year Cost", y = NULL, fill = NULL,
    caption = "Source: RDWG Roadmap, December 2022"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ucla_dark_blue, face = "bold", size = 13),
    plot.caption = element_text(color = "#888888", size = 8),
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

---

## Three Years Have Passed

**None of the four core recommendations have been implemented.**

Since 2022 the environment has intensified:

- NIH DMSP in full effect (January 2023)
- OSTP public access mandate in effect (December 2025)
- Google Drive crisis: **14PB of UCLA research data deleted or migrated** over two years — campus now moving to paid commercial storage (April 2026 deadline active)
- Generative AI created new GPU compute equity issues across every discipline
- Federal funding more competitive and uncertain

**The gaps identified in 2022 are larger now.**

::: {.callout-note appearance="minimal"}
Yale and Cornell document a "dual-change trap": reorganizing staff and building new infrastructure simultaneously produces turmoil. The One IT Design phase (July 2026–June 2027) is where staffing decisions land. This WG's May 15 recommendations should explicitly inform that phase — not arrive after it starts.
:::

---

## What Changed Since 2022

```{mermaid}
quadrantChart
    title Urgency × Anticipation (2026 view)
    x-axis "Roadmap anticipated" --> "Not anticipated"
    y-axis "Lower urgency" --> "Higher urgency"
    quadrant-1 Act now
    quadrant-2 Already planned — still waiting
    quadrant-3 Monitor
    quadrant-4 Emerging priorities
    ORDR not created: [0.12, 0.92]
    P3/P4 enclave missing: [0.18, 0.82]
    Compliance intensification: [0.30, 0.88]
    Federal funding uncertainty: [0.72, 0.78]
    Generative AI: [0.88, 0.90]
    Cloud complexity: [0.38, 0.52]
    Research Software Engineering: [0.48, 0.58]
    Data sovereignty: [0.62, 0.42]
```

---

## What This Group Can Do

| WG Charge Area | RDWG Evidence Available |
|---|---|
| Research data management and storage | Peer benchmarking, Bruin Hub specs, use cases |
| Compliance and regulatory needs | NIH/NSF/OSTP analysis, P3/P4 requirements |
| Compute and infrastructure | Hoffman2 data, cloud analysis |
| Gaps and investment priorities | 8/13 benchmarking gaps, cost analysis |
| Central vs. local IT roles | Full layered model analysis |
| Equity in research access | DSC expansion plan, outreach program |

---

## What I'm Asking

1. **Build on what exists.** 18 months of peer-reviewed evidence is available. We don't need to start from scratch.

2. **Adopt the layered model.** Central coordination augments local expertise — it does not absorb it. This is what works at peer institutions.

3. **Establish clear governance for research data.** Research data is a research function. Coordination should be under VCR, with IT as the infrastructure partner.

4. **Name the unfinished gaps.** ORDR, P3/P4 enclave, baseline storage — three years overdue. This group can give them institutional momentum.

---

## Source Materials

All documents available for download at **[ucla-data-science-center.github.io/rdwg-research-data-infrastructure](https://ucla-data-science-center.github.io/rdwg-research-data-infrastructure/resources.html)**

- **Primary roadmap** — *A Call to Action* (December 2022, 66 pages)
- **Campus engagement report** — *Roadmap Update and Moving Forward* (August 2023, 33 meetings)
- **Peer benchmarking** — *Research on Data Services II* (13 categories)
- **Phase 2 materials** — Implementation task analysis, Deans Council brief, Bruin Data Hub specs

*Tim Dennis, UCLA Library Data Science Center — [tdennis@library.ucla.edu](mailto:tdennis@library.ucla.edu)*
